---
title: "DHFP-Project"
author: "Abdalla Mohamed"
Date: "15/02/2026"
format: html
---

Project Goal: A try to find the factors that help company to be operating / closed / acquired stage.

The Dataset: Information about startup companies and investment via Crunchbase. The investments series at a company level are given. The company status are given. The main dataset is 54294 rows * 39 columns. The demographics will be merged from different databases to find unexpected patterns.

```{r, Reading data}
investments_VC <- read.csv("raw_data_large/investments_VC.csv")

df = data.frame(investments_VC)

summary(df)
```
By checking the columns, it is obvious it needs cleaning. So, in the next step, I am going to clean the dataset from 4856 NA and empty rows depending on funding rounds column, then I will continue exploration.

```{r, Exploration}
df_invest <- df[!is.na(df$funding_rounds) & df$funding_rounds != "", ]

# Calling the save_csv function to save the cleaned file (I asked AI 
# how to call the function from a folder) 
# but later, I find it in lectures
source("Functions/save_csv.R")

# Saving the file
save_csv(df_invest, "raw_data_large/df_invest.csv")

table(df_invest$status)

table(df_invest$funding_rounds)
```

To know the full name of countries to improve readability of the analysis, I downloaded another table but in xlsx, so I tried to read with read.table or read.csv but it didn't work, so I am using readxl library I found according to external google search, it is mentioned also in P. 204 in R material after checking its documentation.

```{r, integration of countries names}
if(!requireNamespace("readxl", quietly=TRUE)) install.packages("readxl")
#?read_excel()

# ROU is the current official ISO 3166-1 alpha-3 country code for Romania. 
# It replaced the former code, ROM, in February 2002 to avoid confusion with
# the Roma people. ROU is also used by the International Olympic Committee
# (IOC). ROM is now considered a transitionally reserved, deprecated code. 
# So, the first step, replacing "ROM" code with "ROU"
# Also, add NA instead of empty cells to avoid unambiguous unmatches
df_invest$country_code[df_invest$country_code == "ROM"] <- "ROU"
df_invest$country_code[df_invest$country_code == ""] <- NA

# Reading the complementary excel file and rename its columns to 
# improve readability
Countries <- readxl::read_excel("Raw_complementary_dataset/3Letters_Countries.xlsx", col_names = c("country_name", "2L_code", "3L_code", "country_code"))

# Merging the two tables
df_invest <- merge(df_invest, Countries[, c("country_name", "3L_code")], 
                   by.x = "country_code", by.y = "3L_code", all.x = TRUE)

# Making a copy cp to add more demographics safely
df_invest_cp <- df_invest

# Saving the resulted merged dataset as df_invest_cp
save_csv(df_invest, "Cleaned_dataset/df_invest_cp.csv")
```
After adding the countries, I realized that there could be ready-made libraries that already do this task, so I installed and used them in the next steps to add more demographic information for the project. But I didn't remove the last merged dataset to keep implementing the concept of integrating multiple datasets.

```{r, Adding the region and the continent}
if(!requireNamespace("countrycode", quietly=TRUE)) install.packages("countrycode")
#?countrycode()

# I am using countrycode specifically because it contains the region
df_invest_cp$Continent <- countrycode::countrycode(sourcevar = df_invest_cp$country_code, origin = "iso3c", destination = "continent")

df_invest_cp$Region <- countrycode::countrycode(sourcevar = df_invest_cp$country_code, origin = "iso3c", destination = "region")
```

I still need more demographics data for comparisons. I need regions like "Middle East", population, GDP per capita and the economics status, so I searched, and found World Development Indicators (World Bank) in WDI library.
```{r, Mergin World Development Indicators (World Bank)}
if(!requireNamespace("WDI", quietly=TRUE)) install.packages("WDI")
#?WDI

World_Demographics <- WDI::WDI(country = "all", indicator = c(
    "gdpPercap" = "NY.GDP.PCAP.KD", "pop" = "SP.POP.TOTL", "lifeExp" = "SP.DYN.LE00.IN"), start = 2023, end = 2026, extra = TRUE)

save_csv(World_Demographics, "Merged_data/World_Demographics.csv")

# By checking the data, 2023 is the last year even it is updated in 2026
World_Demographics23 <- World_Demographics[World_Demographics$year == 2023, ] 

# Remove aggregate regions and NA values
World_Demographics23 <- World_Demographics23[World_Demographics23$region != "Aggregates" & !is.na(World_Demographics23$region), ]

save_csv(World_Demographics23, "Merged_data/World_Demographics23.csv")

# Merging the demographic columns from WDI to the cleaned df_invest_cp 
# with inner join to ensure that all the data now has demographics but 
# I need to remove region column first from df_invest_cp
df_invest_cp <- df_invest_cp[,-10]
df_invest_cp <- merge(x = df_invest_cp, y = 
                        World_Demographics23[, c("iso3c", "gdpPercap", 
                                                 "pop", "lifeExp", 
                                                 "capital", "longitude",
                                                 "latitude", "income")],
                      by.x = "country_code", by.y = "iso3c", all.x = FALSE)

save_csv(df_invest_cp, "Merged_data/df_invest_fullmerged.csv")
```

```{r, 2nd phase of exploring demographics and cleaning the data}
str(df_invest_cp)

# 2nd phase of Cleaning by converting some columns wrong datatypes 
# like funding_total_usd by removing commas and spaces, then converting
# to numeric (I found that R turns "-" into NA automatically,
# and I used this.)
df_invest_cp$funding_total_usd <- gsub("[, ]", "", df_invest_cp$funding_total_usd)
df_invest_cp$funding_total_usd <- as.numeric(df_invest_cp$funding_total_usd)

# Converting coordinates to numeric
df_invest_cp$longitude <- as.numeric(df_invest_cp$longitude)
df_invest_cp$latitude  <- as.numeric(df_invest_cp$latitude)

# Converting all dates using as.Date with specific format in the month
# to get the month name. This format will be useful later in 
# visualizations to show the months names.
df_invest_cp$founded_at       <- as.Date(df_invest_cp$founded_at)
df_invest_cp$first_funding_at <- as.Date(df_invest_cp$first_funding_at)
df_invest_cp$last_funding_at  <- as.Date(df_invest_cp$last_funding_at)
df_invest_cp$month_name <- format(df_invest_cp$founded_at, "%B")

# Factorizing the quarter for easier visualization
df_invest_cp$founded_quarter <- as.factor(
  substr(df_invest_cp$founded_quarter, 6, 7))

save_csv(df_invest_cp, "Merged_data/df_invest_fm_c.csv")
```
Now, I want to know how are startups distribute across Continents?

```{r, Startups distribution accross continents visualizations}
# A place to start from to avoid running the whole code over and over
df_invest_cp <- read.csv("Merged_data/df_invest_fm_c.csv")

source("Functions/visual.R")

# Answering the startups distribution question
par1 = par(mgp = c(4, 1, 0), mar = c(3, 6, 4, 2))

visual(type = "bar", x = sort(table(df_invest_cp$Continent), decreasing = TRUE), title = "Startups by Continent", ylab = "Number of Startups")
```
It is obvious that Americas are the continent of the startups followed by Europe, and ended by Africa.

But, is there a specific month for starting a business? is there a hidden pattern regardless the reasons?

```{r, Visualizing the distribution of Startups foundation by Month}
# Ordering the months first
df_invest_cp$month_name <- factor(df_invest_cp$month_name, levels = month.name)

{
  par2 = par(mgp = c(4, 1, 0), mar = c(7, 6, 4, 2))
  visual(type = "bar", x = table(df_invest_cp$month_name), 
       title = "Number of Founded Startups", xlab = "Month Name", 
       ylab = "Total Foundations", las = 2)
}
```
Results shows that January is weirdly the month of founding a startup! May be some startups consider it representing the whole year.

But does this mean that Q1 will be the highest quarter of founding a startup?
```{r, Visualizing the number of startups founded by Quarter}
# To ensure that missing values has a value in the chart
df_invest_cp$founded_quarter[df_invest_cp$founded_quarter == "" | is.na(df_invest_cp$founded_quarter)] <- "Missing"

visual(type = "bar", x = table(df_invest_cp$founded_quarter),
  title = "Number of Startups Founded by Quarter", xlab = "Quarter",
  ylab = "Count of Startups")
```
Yes, as expected, Q1 is the highest, but this graph shows an opportunity to fill the missing 8701 points as we have the month name then I will check it again.

```{r, Filling the missings in the founded_quarter}
# Filling missings in founded_quarter column with values
quarters <- setNames(rep(paste0("Q", 1:4), each=3), month.name)
df_invest_cp$founded_quarter <- quarters[df_invest_cp$month_name]

# I tried nested ifelse and it worked, may be not the best way, but it is readable.  
# Q1 <- c("January", "February", "March")
# Q2 <- c("April", "May", "June")
# Q3 <- c("July", "August", "September")
# Q4 <- c("October", "November", "December")

# ifelse(df_invest_cp$month_name %in% Q1, "Q1", ifelse(df_invest_cp$month_name %in% Q2, "Q2", ifelse(df_invest_cp$month_name %in% Q3, "Q3", ifelse(df_invest_cp$month_name %in% Q4, "Q4", NA))))

# After material revision of SWITCH rules I found it non vectorized, and only work with sapply, I tried writing it again:
# df_invest_cp$founded_quarter <- sapply(df_invest_cp$month_name, function(x)
#  {switch(x,"January" = "Q1", "February" = "Q1", "March" = "Q1",
#          "April" = "Q2", "May" = "Q2", "June" = "Q2",
#          "July" = "Q3", "August" = "Q3", "September" = "Q3",
#          "October" = "Q4", "November" = "Q4", "December" = "Q4", NA)})
```

After rerunning the plotting function again, it shows clearly that all the missing quarters are filled now and Q1 is still the main quarter.

Let's check the most common status of companies in this dataset?

```{r, Visualizing the current companies status}

df_invest_cp$status[df_invest_cp$status == "" | is.na(df_invest_cp$status)] <- "Missing"

visual(type = "bar", x = sort(table(df_invest_cp$status), decreasing = TRUE),
       title = "Current Companies Status", xlab = "Status", ylab = "Count")
```
Most of the startups are "operating" but there are more than 1108 companies without a status also. I thought removing them will unnecesserily decrease the dataset as these startups could represent some marginalized regions.

So, let's find if the number of funding rounds increase with a country's GDP or not to understand the economic impact!

```{r, Visualizing Country Wealth vs. Investment Rounds}
visual(type = "scatter", x = df_invest_cp$gdpPercap, 
       y = df_invest_cp$funding_rounds, 
       title = "Country Wealth vs. Investment Rounds", 
       xlab = "GDP per Capita", ylab = "Number of Funding Rounds",
       pch = 16, col = adjustcolor("red", alpha.f = 0.1))
```

Yes, it is affecting the low GDP countries, but also affects the high GDP countries. I think that high GDP countries citizens aren't very interested in filling the requirements of the usual funding rounds of startups or may be the higher level of regulations hinder them, or may be got acquired after less number of rounds. The explanation needs more investigation from my opinion to find the confounding factor.

Let's find how many funding rounds do companies usually go through?

```{r, Visualizing Percentage of companies per funding rounds}
# After thinking for some time, this visualization needs 
# the calculation of the cumulative sum
counts <- table(df_invest_cp$funding_rounds)
total_startups <- sum(counts)

# Calculate how many dropped out at each stage
dropped_out <- cumsum(counts)

# Subtracting the total startups number from each round acc. sum
# then I added the original counts (frequencies per funding round)
cum_rounds <- (total_startups - dropped_out) + counts

visual(type = "bar", x = round((cum_rounds / sum(!is.na(df_invest_cp$funding_rounds))) * 100, 1),
  title = "Percentage of companies per Funding Rounds", xlab = "Number of Rounds",
  ylab  = "Percentage of Companies (%)",
  col = "yellow")
```
It is clear that almost the third of the startups that got their first round, will go through a second one, but this gap narrowed after that from the third to almost the half. So, we can call the second round a statistical bottleneck for more rounds.

```{r, Visualizing the mean of different income groups}
# Calculate mean funding by income group
mean_funding <- round(tapply(df_invest_cp$funding_total_usd, df_invest_cp$income, mean, na.rm = TRUE)/100000, 0)


# Plotting the means, a par added here to fit this specific graph
{
par(mgp = c(4, 1, 0), mar = c(10, 6, 4, 2))
  visual(type = "bar", 
         x = mean_funding, ylab = "Mean funding by 100K",
         title = "Mean Funding by Income Group", las = 2)
}
```
It is misleading to get conclusion from this graph only, but knowing the population of these economic categories would reveal a lot.

```{r, Checking the contradiction behind startups funding to economy}
# Calculate the mean funding for each category
mean_funding <- aggregate(funding_total_usd ~ income, data = df_invest_cp, 
                          FUN = mean)

# Get the population for each category (assuming it is a unique value per category)
pop_by_income <- aggregate(pop ~ income, 
                           data = unique(df_invest_cp[, c("country_code",
                                                          "income", "pop")]), 
                           FUN = sum)

# Merging both tables and divide
final_result <- merge(mean_funding, pop_by_income, by = "income")
final_result$fund_access <- round(final_result$funding_total_usd / final_result$pop, 4)
final_result <- final_result[order(final_result$fund_access, decreasing = TRUE), ]

{
  par(mgp = c(4, 1, 0), mar = c(10, 6, 4, 2))
  visual(type = "bar", x = sort(tapply(final_result$fund_access,
                                  final_result$income, unique), 
                                decreasing = TRUE), 
         ylab = "Fund accessibility for startups", 
         title = "Startups Funding Access Per Economy", las = 2)
}
```
This ensures that the accessibility to fund to High income countries is higher than the low income countries according regarding population. It is the douplicated in "High income" countries than "Low income" countries.

Now, I want to predict the success of the Startups. The success here means the startup got acquired by another company. The failure here is closed startup.

```{r, Preparing data for model}
cols_to_keep <- c("status", "funding_total_usd", "funding_rounds", 
                  "founded_year", "seed", "venture", "angel", 
                  "income", "Region")

df_model <- df_invest_cp[, cols_to_keep]
```

```{r, 2nd phase of preparation of data for model}
# Creation of a subset of only 'acquired' and 'closed'
df_exited <- df_model[df_model$status %in% c("acquired", "closed"), ]

# Defining the target: 
# 1 for Success (Acquired)
# 0 for Failure (Closed)
df_exited$is_success <- ifelse(df_exited$status == "acquired", 1, 0)

# Cleaning up any other levels in the factor
df_exited$status <- factor(df_exited$status)
```

```{r, Building the model}
set.seed(123)

# Build the model
exit_model <- glm(is_success ~ funding_total_usd + funding_rounds + 
                    founded_year + seed + venture + angel + income + Region, 
                  family = "binomial", data = df_exited)

summary(exit_model)
```
Each additional round of funding significantly increases the chances of being acquired. This means that the continuous successful funding rounds is an indicator for successful exit (make sense).

While total funding is neutral, specifically having Venture funding has a positive, significant impact.

Being based in North America shows a strong positive correlation with acquisition.

Founded Year coefficient is highly negative. This means, as the year goes up (newer companies), the likelihood of having been acquired goes down. This makes sense also as older companies have had more time to exit.

Surprisingly, higher seed funding has a tiny but statistically significant negative impact on the probability of acquisition in this model. This could imply that companies that highly depend on seed without moving to venture are less likely to exit.

```{r, Producing the matrix of predictions}
exit_preds <- predict(exit_model, df_exited, type = "response")
exit_class <- ifelse(exit_preds > 0.5, 1, 0)
table(Predicted = exit_class, Actual = df_exited$is_success)
```

```{r, Calculating the accuracy}
conf_matrix <- table(Predicted = exit_class, Actual = df_exited$is_success)

# Calculating Accuracy
# Formula: (True Positives + True Negatives) / Total
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
paste0("Accuracy of the model is: ", round(accuracy * 100, 2), "%")
```

```{r, Visualizing the comparison between Acquired vs Closed}
# Comparison of the predicted probabilities of the two groups
visual(type = "box", x = exit_preds, y = df_exited$is_success, range = 0,
        title = "Predicted Probability: Acquired (1) vs Closed (0)",
        xlab = "0=Closed, 1=Acquired", 
        ylab = "Probability of Success",
        col = c("red", "green"))

```

```{r, Checking the acquired companies but wrongly predicted}
# Add the predictions back to the filtered 'exited' dataframe
df_exited$predicted_class <- exit_class

# Identify the 'False Negatives' (Actual: Acquired, Predicted: Closed)
# These are the 891 companies
surprising_acquisitions_data <- df_exited[df_exited$is_success == 1 & df_exited$predicted_class == 0, ]

# Pulling the names from the original dataframe using the row indices
# I used the company name column to identify them
surprising_names <- df_invest_cp[rownames(surprising_acquisitions_data), c("name", "market", "funding_total_usd", "funding_rounds", "country_name", "Region")]

# Saving the first 20 Markets of unpredicted successful companies
surprising_names20 <- head(sort(table(surprising_names$market), 
                                decreasing = TRUE), 20)

# Cleaning all the names of unpredicted successful companies
surprising_names <- na.omit(surprising_names)
```

```{r, Visualizing the Top 20 markets for unpredicted successful startups}
{
  par(mar = c(10.5, 4, 2, 2))
  visual(type = "bar", surprising_names20, las = 2, col = "green", 
                title = "Top 20 Markets for Successful Acquisitions",
                ylab = "Number of Companies")
}
```
Checking the possible reasons behind unprediction

```{r, Checking the possible reasons behind unpredictability}
sapply(list(original = df_invest_cp[df_invest_cp$status == "acquired", ], 
            unpredicte = surprising_names),
       function(df) {
         c(Mean_Funding_100K = mean(df$funding_total_usd / 100000, na.rm = TRUE),
           Mean_Rounds = mean(df$funding_rounds, na.rm = TRUE))
       })
```
The difference between means require more investigation, so let's test the significance of this difference.

```{r, Check if the difference in means is significant or not}
# Defining the two comparison groups
group_original <- df_invest_cp[df_invest_cp$status == "acquired", ]
group_surprising <- surprising_names

# Defining the columns I want to test
cols_to_test <- c("funding_total_usd", "funding_rounds")

# Running the t-test for each column
t_results <- sapply(cols_to_test, function(col) {
  test <- t.test(group_original[[col]], group_surprising[[col]], na.rm = TRUE)
  c(p_value = test$p.value, 
  t_stat = round(test$statistic, 5),
  significant = ifelse(test$p.value < 0.05, "YES", "NO"))
  })

# Display the results
print(t_results)
```

According to t.test result, the difference between the means of these two columns in the unpredicted dataset and the whole dataset, shows statistical significance. This explain the unpredictability of these companies as a successful companies by the model. Note, I just tested the numeric columns, the other columns like "country", "Region" and "market", could have their weight too.