---
title: "DHFP-Project"
author: "Abdalla Mohamed"
Date: "15/02/2026"
format: html
---

Project Goal: A try to understand some patterns that help company to be closed / acquired.

The Dataset: Information about startup companies and investment via Crunchbase. The investments series at a company level are given. The company status are given. The main dataset is 54294 rows * 39 columns. The demographics will be merged from different databases to find unexpected patterns.

```{r, Reading data}
investments_VC <- read.csv("raw_data_large/investments_VC.csv")

df_invest = data.frame(investments_VC)

summary(df_invest)

# Calling the save_csv function to save the cleaned file 
# (I asked AI how to call the function from a folder) 
# but later, I found it in lectures
source("Functions/save_csv.R")

# Saving the file
save_csv(df_invest, "raw_data_large/df_invest.csv")
```
By checking the columns summary, it is obviously needs cleaning. So, in the next step, I am going to clean the dataset from 4856 NA and empty rows depending on "funding rounds" column as one of the main columns this analysis will be built on, then I will continue exploration for further cleaning.

```{r, 1st phase of cleaning followed by Exploration}
df_invest <- df_invest[!is.na(df_invest$funding_rounds) 
                       & df_invest$funding_rounds != "", ]

# Saving the file using my customized function save_csv in Functions folder
save_csv(df_invest, "Cleaned_dataset/df_invest.csv")

# Continue the exploration
table(df_invest$status)

table(df_invest$funding_rounds)

```

To know the full name of countries to improve readability of the analysis, I downloaded another table but in xlsx, so I tried to read with read.table or read.csv but it didn't work, so I am using readxl library I found according to external google search, it is mentioned also in P. 204 in R material after checking its documentation.

```{r, Integration of countries names}
if(!requireNamespace("readxl", quietly=TRUE)) install.packages("readxl")
#?read_excel()

# ROU is the current official ISO 3166-1 alpha-3 country code for Romania. 
# It replaced the former code, ROM. 
# ROU is also used by the International Olympic Committee (IOC).
# So, the first step, replacing "ROM" code with "ROU"
# Also, I am adding NA instead of "" cells to avoid unambiguous mismatches
df_invest$country_code[df_invest$country_code == "ROM"] <- "ROU"
df_invest$country_code[df_invest$country_code == ""] <- NA

# Reading the complementary excel file and rename its columns to 
# improve readability
Countries <- readxl::read_excel("Raw_complementary_dataset/3Letters_Countries.xlsx", col_names = c("country_name", "2L_code", "3L_code", "country_code"))

# Merging the two tables (excel file + my cleaned file)
df_invest <- merge(df_invest, Countries[, c("country_name", "3L_code")], 
                   by.x = "country_code", by.y = "3L_code", all.x = TRUE)

# Saving the resulted merged dataset as df_invest_cp
save_csv(df_invest, "Cleaned_dataset/df_invest_cp.csv")

# Making a copy (cp) to add more demographics safely
df_invest_cp <- df_invest
```
After adding the countries names, I realized that there could be ready-made libraries that already useful to do this task, so I installed and used one of them in the next step to add more demographic information for the project. But I didn't remove the last merged dataset to keep implementing the concept of integrating multiple datasets.

```{r, Adding the region and the continent}
if(!requireNamespace("countrycode", quietly=TRUE)) install.packages("countrycode")
#?countrycode()

# I am using countrycode specifically because it contains the region
df_invest_cp$Continent <- countrycode::countrycode(sourcevar = df_invest_cp$country_code, origin = "iso3c", destination = "continent")

df_invest_cp$Region <- countrycode::countrycode(sourcevar = df_invest_cp$country_code, origin = "iso3c", destination = "region")
```

I still need more demographics data for possible comparisons. I need population, GDP per capita and the economics status category, so I searched, and found World Development Indicators (World Bank) in WDI library.
```{r, Mergin World Development Indicators (World Bank)}
if(!requireNamespace("WDI", quietly=TRUE)) install.packages("WDI")
#?WDI

World_Demographics <- WDI::WDI(country = "all", indicator = c(
    "gdpPercap" = "NY.GDP.PCAP.KD", "pop" = "SP.POP.TOTL", "lifeExp" = "SP.DYN.LE00.IN"), start = 2023, end = 2026, extra = TRUE)

save_csv(World_Demographics, "Raw_complementary_dataset/World_Demographics.csv")

# By checking the data, 2023 is the last year even it is updated in 2026
World_Demographics23 <- World_Demographics[World_Demographics$year == 2023, ] 

# Remove aggregate regions and NA values
World_Demographics23 <- World_Demographics23[World_Demographics23$region != "Aggregates" & !is.na(World_Demographics23$region), ]

save_csv(World_Demographics23, "Cleaned_dataset/World_Demographics23.csv")

# Merging the demographic columns from WDI to the cleaned df_invest_cp 
# with inner join to ensure that all the cleaned data now has demographics 
# but I want to remove region column first from df_invest_cp 
df_invest_cp <- df_invest_cp[,-10]

# all.x = FALSE is the default, but I let it to confirm inner join
# as it will decrease the rows of df_invest_cp intentionally to 44122
df_invest_cp <- merge(x = df_invest_cp, y = 
                        World_Demographics23[, c("iso3c", "gdpPercap", 
                                                 "pop", "lifeExp", 
                                                 "capital", "longitude",
                                                 "latitude", "income")],
                      by.x = "country_code", by.y = "iso3c", all.x = FALSE)

save_csv(df_invest_cp, "Merged_data/df_invest_fullmerged.csv")
```

```{r, 2nd phase of exploring demographics and cleaning the data}
str(df_invest_cp)

# 2nd phase of Cleaning by converting some columns wrong datatypes 
# like funding_total_usd by removing commas and spaces, then converting
# to numeric (I realized that R turns "-" into NA automatically,
# and I used it)
df_invest_cp$funding_total_usd <- gsub("[, ]", "", df_invest_cp$funding_total_usd)
df_invest_cp$funding_total_usd <- as.numeric(df_invest_cp$funding_total_usd)

# Converting coordinates to numeric
df_invest_cp$longitude <- as.numeric(df_invest_cp$longitude)
df_invest_cp$latitude  <- as.numeric(df_invest_cp$latitude)

# Converting all dates using as.Date with specific format in the month
# to get the month name. This format will be useful later in 
# visualizations to show the months names.
df_invest_cp$founded_at       <- as.Date(df_invest_cp$founded_at)
df_invest_cp$first_funding_at <- as.Date(df_invest_cp$first_funding_at)
df_invest_cp$last_funding_at  <- as.Date(df_invest_cp$last_funding_at)
df_invest_cp$month_name <- format(df_invest_cp$founded_at, "%B")

# Factorizing the quarter for easier visualization
df_invest_cp$founded_quarter <- 
  as.factor(substr(df_invest_cp$founded_quarter, 6, 7))

save_csv(df_invest_cp, "Merged_data/df_invest_fm_c.csv")
```
Now, I want to know how are startups distribute across Continents?

```{r, Startups distribution accross continents visualizations}
# A place to start from, to avoid running the whole code every time
df_invest_cp <- read.csv("Merged_data/df_invest_fm_c.csv")

# Calling the visual function customized for plotting and writing text 
source("Functions/visual.R")

# Answering the startups distribution across continents question
par(mgp = c(4, 1, 0), mar = c(3, 6, 4, 2))

visual(type = "bar", x = sort(table(df_invest_cp$Continent), decreasing = TRUE), title = "Startups by Continent", ylab = "Number of Startups")
```
It is obvious that Americas are the continent of the startups followed by Europe, and ended by Africa.

But, is there a specific month for starting a business? is there a hidden pattern regardless the reasons?

```{r, Visualizing the distribution of Startups foundation by Month}
# Ordering the months first and I used the month.name function here :)
df_invest_cp$month_name <- factor(df_invest_cp$month_name,
                                  levels = month.name)

{
  par(mgp = c(4, 1, 0), mar = c(7, 6, 4, 2))
  visual(type = "bar", x = table(df_invest_cp$month_name), 
       title = "Number of Founded Startups", xlab = "Month Name", 
       ylab = "Total Foundations", las = 2)
}
```
Results shows that January is weirdly the month of founding a startup! May be some startups consider it representing the whole year.

But does this mean that Q1 will be the highest quarter of founding a startup?
```{r, Visualizing the number of startups founded by Quarter}
# To ensure that missing values has a value in the chart
df_invest_cp$founded_quarter[df_invest_cp$founded_quarter == "" | is.na(df_invest_cp$founded_quarter)] <- "Missing"


visual(type = "bar", x = table(df_invest_cp$founded_quarter),
  title = "Number of Startups Founded by Quarter", xlab = "Quarter",
  ylab = "Count of Startups")
```
Yes, as expected, Q1 is the highest, but this graph shows an opportunity to fill the missing 8701 points as we have the month name then I will check it again.

```{r, Filling the missings in the founded_quarter}
# Filling Missings in founded_quarter column with values
quarters <- setNames(rep(paste0("Q", 1:4), each=3), month.name)
df_invest_cp$founded_quarter <- quarters[df_invest_cp$month_name]

# I tried nested ifelse and it worked, may be not the best way, but it is readable.  
# Q1 <- c("January", "February", "March")
# Q2 <- c("April", "May", "June")
# Q3 <- c("July", "August", "September")
# Q4 <- c("October", "November", "December")

# ifelse(df_invest_cp$month_name %in% Q1, "Q1", ifelse(df_invest_cp$month_name %in% Q2, "Q2", ifelse(df_invest_cp$month_name %in% Q3, "Q3", ifelse(df_invest_cp$month_name %in% Q4, "Q4", NA))))

# After material revision of SWITCH rules I found it non vectorized, and only work with sapply, I tried writing it again:
# df_invest_cp$founded_quarter <- sapply(df_invest_cp$month_name, function(x)
#  {switch(x,"January" = "Q1", "February" = "Q1", "March" = "Q1",
#          "April" = "Q2", "May" = "Q2", "June" = "Q2",
#          "July" = "Q3", "August" = "Q3", "September" = "Q3",
#          "October" = "Q4", "November" = "Q4", "December" = "Q4", NA)})
```

After rerunning the plotting function again, it shows clearly that all the missing quarters are filled now and Q1 is still the main quarter.

Let's check the most common status of companies in this dataset?

```{r, Visualizing the current companies status}

df_invest_cp$status[df_invest_cp$status == "" | is.na(df_invest_cp$status)] <- "Missing"

visual(type = "bar", x = sort(table(df_invest_cp$status), decreasing = TRUE),
       title = "Current Companies Status", xlab = "Status", ylab = "Count")
```
Most of the startups are "operating" but there are more than 1108 companies without a status also. I thought removing them will unnecesserily decrease the dataset as these startups could represent some marginalized regions.

So, let's find if the number of funding rounds increase with a country's GDP or not to understand the economic impact!

```{r, Visualizing Country Wealth vs. Investment Rounds}
visual(type = "scatter", x = df_invest_cp$gdpPercap, 
       y = df_invest_cp$funding_rounds, 
       title = "Country Wealth vs. Investment Rounds", 
       xlab = "GDP per Capita", ylab = "Number of Funding Rounds",
       pch = 16, col = adjustcolor("red", alpha.f = 0.1))
```

Yes, it is affecting the low GDP countries, but also affects the high GDP countries. I think that high GDP countries citizens might not very interested in filling the requirements of the usual funding rounds of startups, or may be the higher level of regulations hinder them, or may be got acquired instantly without going to funding rounds. Anyway, there is a clear gap, and the explanation needs more investigation from my opinion to find the confounding factors.

Let's find how many funding rounds do companies usually go through?

```{r, Visualizing Percentage of companies per funding rounds}
# After thinking for some time, this visualization needs 
# the calculation of the cumulative sum as the startup that funded for 18 
# rounds must have passed through the first, second till the 17th rounds
counts <- table(df_invest_cp$funding_rounds)
total_startups <- sum(counts)

# Calculate how many dropped out at each stage
dropped_out <- cumsum(counts)

# Subtracting the total startups number from each round acc. sum
# then I added the original counts (frequencies per funding round)
cum_rounds <- (total_startups - dropped_out) + counts

visual(type = "bar", x = round((cum_rounds / sum(!is.na(df_invest_cp$funding_rounds))) * 100, 1),
  title = "Percentage of companies per Funding Rounds", xlab = "Number of Rounds",
  ylab  = "Percentage of Companies (%)",
  col = "yellow")
```
Almost the third of the startups that got their first round, will go through a second one, but this gap narrowed after that from the third to the half. So, we can call the second round a statistical bottleneck for more rounds.

```{r, Visualizing the mean of different income groups}
# Calculating the mean funding by income group then added as the x
# Plotting the means, a par added here to fit this specific graph
{
  par(mgp = c(4, 1, 0), mar = c(10, 6, 4, 2))
  visual(type = "bar", 
         x = round(tapply(df_invest_cp$funding_total_usd, df_invest_cp$income,
                          mean, na.rm = TRUE)/100000, 0),
         ylab = "Mean funding by 100K",
         title = "Mean Funding by Income Group", las = 2)
}
```
It is misleading to get conclusions from this graph or GDP graph directly, but knowing the population of these economic categories might reveal different conclusions.

```{r, Checking the contradiction behind startups funding to economy}
# Calculate the mean funding for each category
mean_funding <- aggregate(funding_total_usd ~ income, data = df_invest_cp, 
                          FUN = mean)

# Get the population for each category (assuming it is a unique value per category)
pop_by_income <- aggregate(pop ~ income, 
                           data = unique(df_invest_cp[, c("country_code",
                                                          "income", "pop")]), 
                           FUN = sum)

# Merging both tables and divide
final_result <- merge(mean_funding, pop_by_income, by = "income")
final_result$fund_access <- round(final_result$funding_total_usd / final_result$pop, 4)
final_result <- final_result[order(final_result$fund_access, decreasing = TRUE), ]

{
  # Setting the layout (1 row, 2 columns)
  par(mfrow = c(1, 2), mgp = c(4, 1, 0), mar = c(10, 6, 4, 2))
  
  # Plotting the Population by Income
  visual(type = "bar", sort(round(tapply(final_result$pop,
                                         final_result$income, sum) / 
                                    100000000, 1), decreasing = TRUE), 
         ylab = "Total Population in 100M", 
         title = "Population per Income Group", las = 2)
  
  # Plotting the Funding Access
  visual(type = "bar", x = sort(tapply(final_result$fund_access, 
                           final_result$income, unique), 
                    decreasing = TRUE), 
         ylab = "Fund accessibility for startups", 
         title = "Startups Funding Access", las = 2)
  
  # Resetting layout to 1x1 so future plots aren't affected
  par(mfrow = c(1, 1))
}
```

These two graphs ensures that the accessibility to fund in High income countries is higher than the low income countries according to population. It is duplicated in "High income" countries than "Low income" countries. Although, the "Lower Middle Income" & "Upper Middle Income" countries have much higher populations, but their access to fund regarding their population is still uneven. This shows the unfair access to fund in the world.

As a digital health master's student, I want to check if the healthcare services are among the top markets acquired or not.

```{r, checking the healthcare services among the top markets acquired}
# Calculating the acquisition rate considering the mean of the logical vector
# Where TRUE (for acquired) becomes 1 and FALSE becomes 0
# So the mean is the percentage
market_rates <- tapply(df_invest_cp$status == "acquired", 
                         df_invest_cp$market, 
                         mean, na.rm = TRUE)
  
# Filtering for markets with at least 10 companies to ensure meaningful 
# interpretation then sorting
top_markets <- sort(market_rates[(table(df_invest_cp$market)) >= 10], decreasing = TRUE)
  
# Visualising the top 10 markets
{
  par(mar = c(10.5, 4, 2, 2))
  visual(type = "bar", head(round(top_markets * 100, 0), 10), 
         las = 2, col = "green", 
         title = "Top 20 Markets for Successful Acquisitions",
         ylab = "Percentage of Mean Companies")
}
```
This shows that Healthcare Services ranked the sixth in the top 

Now, I try to predict the success of Startups. The success here means the startup got acquired by another company. The failure here is "closed" startup. (I used AI to help me with this prediction part, under my supervision.)

```{r, Preparing data for model}
# The columns I decided to use in the prediction where status is y 
# and the rest of columns are predictors
cols_to_keep <- c("status", "funding_total_usd", "funding_rounds", 
                  "founded_year", "seed", "venture", "angel", 
                  "income", "Region")

df_model <- df_invest_cp[, cols_to_keep]
```

```{r, 2nd phase of preparation of data for model}
# Creation of a subset of only 'acquired' and 'closed'
df_exited <- df_model[df_model$status %in% c("acquired", "closed"), ]

# Defining the target: 
# 1 for Success (Acquired)
# 0 for Failure (Closed)
df_exited$is_success <- ifelse(df_exited$status == "acquired", 1, 0)

# Cleaning up any other levels in the factor
df_exited$status <- factor(df_exited$status)
```

```{r, Building the model}
set.seed(123)

# Build the model
exit_model <- glm(is_success ~ funding_total_usd + funding_rounds + 
                    founded_year + seed + venture + angel + income + Region, 
                  family = "binomial", data = df_exited)

summary(exit_model)
```
Each additional round of funding significantly increases the chances of being acquired. This means that the continuous successful funding rounds is an indicator for successful exit (make sense).

While total funding is neutral, specifically having Venture funding has a positive, significant impact.

Being based in North America shows a strong positive correlation with acquisition.

Founded Year coefficient is highly negative. This means, as the year goes up (newer companies), the likelihood of having been acquired goes down. This makes sense also as older companies have had more time to exit and achieve higher revenues and higher potential for investors.

Surprisingly, higher seed funding has a tiny but statistically significant negative impact on the probability of acquisition in this model. This could imply that companies that highly depend on seed without moving to venture are less likely to exit. This makes sense as ventures are more involved in investors' network and have a clear benefit from selling the company.

```{r, Producing the matrix of predictions}
exit_preds <- predict(exit_model, df_exited, type = "response")
exit_class <- ifelse(exit_preds > 0.5, 1, 0)
table(Predicted = exit_class, Actual = df_exited$is_success)
```

```{r, Calculating the accuracy}
conf_matrix <- table(Predicted = exit_class, Actual = df_exited$is_success)

# Calculating Accuracy
# Formula: (True Positives + True Negatives) / Total
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
paste0("Accuracy of the model is: ", round(accuracy * 100, 2), "%")
```

```{r, Visualizing the comparison between Acquired vs Closed}
# Comparison of the predicted probabilities of the two groups
visual(type = "box", x = exit_preds, y = df_exited$is_success, range = 0,
        title = "Predicted Probability: Acquired (1) vs Closed (0)",
        xlab = "0=Closed, 1=Acquired", 
        ylab = "Probability of Success",
        col = c("red", "green"))

```

```{r, Checking the acquired companies but wrongly predicted}
# Add the predictions back to the filtered 'exited' dataframe
df_exited$predicted_class <- exit_class

# Identify the 'False Negatives' (Actual: Acquired, Predicted: Closed)
# These are the 891 companies
surprising_acquisitions_data <- df_exited[df_exited$is_success == 1 & df_exited$predicted_class == 0, ]

# Pulling the names from the original dataframe using the row indices
# I used the company name column to identify them
surprising_names <- df_invest_cp[rownames(surprising_acquisitions_data), c("name", "market", "funding_total_usd", "funding_rounds", "country_name", "Region")]

# Saving the first 20 Markets of unpredicted successful companies
surprising_names20 <- head(sort(table(surprising_names$market), 
                                decreasing = TRUE), 20)

# Cleaning all the names of unpredicted successful companies
surprising_names <- na.omit(surprising_names)
```

```{r, Visualizing the Top 20 markets for unpredicted successful startups}
{
  par(mar = c(10.5, 4, 2, 2))
  visual(type = "bar", surprising_names20, las = 2, col = "green", 
                title = "Top 20 Markets for Successful Acquisitions",
                ylab = "Number of Companies")
}
```
Checking the possible reasons behind unprediction

```{r, Checking the possible reasons behind unpredictability}
sapply(list(original = df_invest_cp[df_invest_cp$status == "acquired", ], 
            unpredicte = surprising_names),
       function(df) {
         c(Mean_Funding_100K = mean(df$funding_total_usd / 100000, na.rm = TRUE),
           Mean_Rounds = mean(df$funding_rounds, na.rm = TRUE))
       })
```
The difference between means require more investigation, so let's test the significance of this difference.

```{r, Check if the difference in means is significant or not}
# Defining the two comparison groups
group_original <- df_invest_cp[df_invest_cp$status == "acquired", ]
group_surprising <- surprising_names

# Defining the columns I want to test
cols_to_test <- c("funding_total_usd", "funding_rounds")

# Running the t-test for each column
t_results <- sapply(cols_to_test, function(col) {
  test <- t.test(group_original[[col]], group_surprising[[col]], na.rm = TRUE)
  c(p_value = test$p.value, 
  t_stat = round(test$statistic, 5),
  significant = ifelse(test$p.value < 0.05, "YES", "NO"))
  })

# Display the results
print(t_results)
```

According to t.test result, the difference between the means of these two columns in the unpredicted dataset and the whole "acquired" dataset, shows statistical significance. This explain the unpredictability of these companies as a successful companies by the model.

Note, I just tested the numeric columns, the other columns like "country", "Region" and "market", could have their weight too.