---
title: "DHFP-Project"
author: "Abdalla Mohamed"
Date: "15/02/2026"
format: html
---

Project Goal: I would like to find the factors that help company to be operating / closed / acquired stage.

The Dataset: Information about startup companies and investment via Crunchbase. The investments series at a company level are given. The company status are given. The main dataset is 54294 rows * 39 columns. The demographics will be merged from different databases to find unexpected patterns.

```{r, Reading data}
investments_VC <- read.csv("F:/Digital Health Master's Potsdam/1st Semester DH/Fundmentals of Programming R/Project/investments_VC.csv")

df = data.frame(investments_VC)

summary(df)
```
# Cleaning the dataset from NA and empty rows depending on funding rounds column, then continue exploration.

```{r, Exploration}
df_invest <- df[!is.na(df$funding_rounds) & df$funding_rounds != "", ]

# Call save_csv function to save the cleaned file 
source("F:/Digital Health Master's Potsdam/1st Semester DH/Fundmentals of Programming R/Project/Functions/save_csv.R")

# Saving the file
save_csv(df_invest, "df_invest.csv")

table(df_invest$status)

table(df_invest$funding_rounds)
```

To know the full name of countries, I downloaded another table but in xlsx, so I tried to read with read.table or read.csv but it didn't work, so I am using readxl library

```{r, integration of countries names}
library("readxl")
#?read_excel()

# Reading the complementary excel file and rename its columns to improve readability
Countries <- read_excel("F:/Digital Health Master's Potsdam/1st Semester DH/Fundmentals of Programming R/Project/3Letters_Countries.xlsx", col_names = c("country_name", "iso_alpha2", "iso_alpha3", "country_code"))

# Merging the two tables
df_invest <- merge(df_invest, Countries[, c("country_name", "iso_alpha3")], by.x = "country_code", by.y = "iso_alpha3", all.x = TRUE)

# Making a copy cp to add more demographics safely
df_invest_cp <- df_invest

# Saving the resulted merged dataset as df_invest_cp
save_csv(df_invest, "df_invest_cp.csv")
```

```{r}
# After adding the countries, I realized that there could be ready-made libraries that already do this task, so I installed and used them in the next steps to add more demographic information for the project.

library(countrycode)
#?countrycode()

# I am using countrycode specifically because it contains the region, gapminder didn't have regions, but I will use it later as it has more demographic features.
df_invest_cp$Continent <- countrycode(sourcevar = df_invest_cp$country_code,
                                        origin = "iso3c",
                                        destination = "continent")

df_invest_cp$Region <- countrycode(sourcevar = df_invest_cp$country_code,
                                     origin = "iso3c",
                                     destination = "region")


# install.packages("WDI")
library(WDI)
#?WDI

World_Demographics <- WDI(country = "all", indicator = c(
    "gdpPercap" = "NY.GDP.PCAP.PP.KD", "pop" = "SP.POP.TOTL", 
    "lifeExp" = "SP.DYN.LE00.IN"), start = 2000, end = 2026, extra = TRUE)

save_csv(World_Demographics, "World_Demographics.csv")

# By checking the data, 2023 is the last year even it is updated in 2026
World_Demographics23 <- World_Demographics[World_Demographics$year == 2023, ] 
# Subsetting aggregate regions (like "Arab World") to keep only countries and NA values
World_Demographics23_agg <- World_Demographics23[World_Demographics23$region == "Aggregates" & !is.na(World_Demographics23$region), ]

save_csv(World_Demographics23_agg, "World_Demographics23_agg.csv")

# Remove aggregate regions and NA values
World_Demographics23 <- World_Demographics23[World_Demographics23$region != "Aggregates" & !is.na(World_Demographics23$region), ]

save_csv(World_Demographics23, "World_Demographics23.csv")

# Merging the demographic columns from WDI to investment_VC1
df_invest_cp <- merge(x = df_invest_cp, y = World_Demographics[, c("iso3c", "gdpPercap", "pop", "lifeExp", "capital", "longitude", "latitude", "income", "lending")], by.x = "country_code", by.y = "iso3c", all.x = FALSE)

save_csv(df_invest_cp, "df_invest_fullmerged.csv")
```

```{r, exploring demographics}
table(df_invest_cp$Region)
```
